### 2025-05-27 docs/article/Sections/introduction.tex

Efficient thread utilization is crucial for low-thread servers. These servers
must handle many concurrent requests without being overwhelmed, which
necessitates asynchronous, non-blocking request handling
~\cite{carvalho2023async}. By leveraging asynchronous I/O operations—such as
database queries and API calls, servers can avoid blocking threads while
waiting for data, thereby maximizing throughput and responsiveness. To support
this non-blocking architecture, PSSR implementations require template engines
that are compatible with asynchronous data models and I/O. Some modern template
engines, such as HtmlFlow~\cite{htmlflow} and Thymeleaf~\cite{thymeleaf}, have
been designed with these capabilities in mind. However, many traditional
template engines—particularly those using external domain-specific languages
~\cite{Fowler03} (DSLs)—still depend on blocking interfaces like
\texttt{Iterable} for data processing. This blocking behavior forces server
threads to remain idle until the entire HTML output is ready, undermining the
performance benefits of non-blocking I/O and limiting scalability in
high-concurrency environments.

With the introduction of \textit{Virtual Threads}~\footnote{\url{https://openjdk.org/jeps/444}}
in Java 21, it is now possible to execute blocking I/O operations in a
scalable, lightweight manner. This capability allows traditional template
engines—often reliant on blocking interfaces—to operate efficiently in
high-concurrency, non-blocking environments without requiring complex
asynchronous programming models. As a result, Progressive Server-Side Rendering
(PSSR) can now be implemented using familiar HTML templates, 
simplifying development and improving maintainability.

This work explores the current landscape of non-blocking PSSR, focusing on two
primary paradigms: \textit{reactive programming} and \textit{coroutines}, both
of which have traditionally been used to achieve asynchronous I/O. As an
alternative, we investigate whether Java’s Virtual Threads can offer comparable
performance while preserving the simplicity of synchronous code. Section 2
reviews the state-of-the-art in PSSR and template engine design. Section 3
outlines the limitations of conventional engines in asynchronous settings and
presents our proposed approach. Section 4 details the benchmark methodology,
followed by the results and analysis in Section 5. Conclusions are presented in Section 6.

>>>>

\textit{Low-thread} servers, also known as
\textit{event-driven}~\cite{event-driven-servers}, have gained prominence in
contemporary web applications due to their ability to efficiently manage a large
number of concurrent I/O operations with minimal resources, thus promoting
better scalability.
By leveraging asynchronous I/O operations—such as
database queries and API calls, servers can avoid blocking threads while
waiting for data, thereby maximizing throughput and responsiveness. To support
this non-blocking architecture, PSSR implementations require template engines
that are compatible with asynchronous data models~\cite{carvalho2023async}.
Some modern template
engines, such as HtmlFlow~\cite{htmlflow} and Thymeleaf~\cite{thymeleaf}, have
been designed with these capabilities in mind. However, many traditional
template engines—particularly those using external domain-specific languages
(DSLs)~\cite{Fowler03}—still depend on blocking interfaces like
\texttt{Iterable} for data processing. This blocking behavior forces server
threads to remain idle until the entire HTML output is ready, undermining the
performance benefits of non-blocking I/O and limiting scalability in
high-concurrency environments.

With the introduction of \textit{Virtual Threads}~\footnote{\url{https://openjdk.org/jeps/444}}
in Java 21, it is now possible to execute blocking I/O operations in a
scalable, lightweight manner. This capability allows traditional template
engines—often reliant on blocking interfaces—to operate efficiently in
high-concurrency, non-blocking environments without requiring complex
asynchronous programming models. As a result, PSSR
can now be implemented using familiar HTML templates, 
simplifying development and improving maintainability.

This work explores the current landscape of non-blocking PSSR, focusing on two
primary paradigms: \textit{reactive programming} and \textit{coroutines}, both
of which have been used to achieve asynchronous I/O in the Java ecosystem. As an
alternative, we investigate whether Java’s Virtual Threads can offer comparable
performance while preserving the simplicity of synchronous code. Section 2
reviews the state-of-the-art in PSSR and template engine design. Section 3
outlines the limitations of conventional engines in asynchronous settings and
presents our proposed approach. Section 4 details the benchmark methodology,
followed by the results and analysis in Section 5. Conclusions are presented in Section 6.

### 2025-05-27 docs/article/Sections/problem-statement.tex

In this section, we examine the challenges of implementing Progressive
Server-Side Rendering (PSSR) in modern web applications, with a focus on the
limitations of current template engine designs. Our goal is to broaden the
range of options available for PSSR, particularly within JVM-based frameworks.

Reactive streams~\cite{ReactiveStreams}, such as those based on
\texttt{Observable} (from RxJava)~\cite{rx-observable} and \texttt{Flow} (from
the Java 9+ standard library~\footnote{~\url{https://openjdk.org/jeps/266}}),
are abstractions that provide a non-blocking, asynchronous
way to handle data streams.
These abstractions support the incremental generation and transmission of HTML
content as data becomes available, enabling \textit{progressive server-side
rendering} (PSSR)~\cite{pssr2005}.
In PSSR, the server does not
wait for the entire data model to be ready before beginning to render HTML.
Instead, it processes and streams each piece of data to the client as soon as
it arrives.

Reactive types like \texttt{Observable<T>} or \texttt{Flow<T>}
facilitate this by representing data as a sequence of asynchronous events.
These streams emit values over time, allowing the
rendering engine to consume data at a manageable rate without overwhelming
system resources.
For example, a reactive stream might emit a sequence of \texttt{Presentation}
objects, each representing a talk in a conference schedule.
As each \texttt{Presentation} is emitted by the \texttt{Observable}
or \texttt{Flow}, the internal DSL-based engine—such as HtmlFlow—can render the
corresponding HTML fragment and immediately flush it to the client. This
approach is demonstrated in \autoref{lst:presentation-observable}, where each
presentation is rendered asynchronously as it is emitted by an
\texttt{Observable}.
Note that the \texttt{await} builder receives an additional parameter, the
\texttt{onCompletion} callback, which is used to signal HtmlFlow that it can
proceed to render the next HTML element in the web
template~\cite{carvalho2023async}.
HtmlFlow pauses the rendering process until \texttt{onCompletion} is called,
similar to how the \emph{resume} function works in continuations and
coroutines~\cite{coroutines_continuations}.
In \autoref{lst:presentation-flow}, we show an
equivalent suspend-based implementation using Kotlin's \texttt{Flow}~\cite{PSSR-WISE2024}.
Both examples highlight how internal DSLs can natively integrate with reactive types
to enable non-blocking, progressive rendering on the server side.

\lstset{style=listingstyle}
\begin{center}
\begin{minipage}{0.50\textwidth}
\begin{lstlisting}[
    language=Kotlin,
    basicstyle=\scriptsize\ttfamily,
    numbers=none,
    caption={\textit{HtmlFlow} presentation template with \texttt{Observable}},
    label={lst:presentation-observable}
]
await { div, model, onCompletion ->
  model
    .doOnNext { presentation ->
      presentationFragmentAsync
        .renderAsync(presentation)
        .thenApply { frag -> div.raw(frag) }
      }
    .doOnComplete { onCompletion.finish() }
    .subscribe()
}
\end{lstlisting}
\end{minipage}
\hfill
\begin{minipage}{0.46\textwidth}
\begin{lstlisting}[
    language=Kotlin,
    basicstyle=\scriptsize\ttfamily,
    numbers=none,
    label={lst:presentation-flow},
    caption={\textit{HtmlFlow} presentation template with \texttt{Flow}},
]
suspending { model ->
  model
    .toFlowable()
    .asFlow()
    .collect { presentation ->
      presentationFragmentAsync
        .renderAsync(presentation)
        .thenApply { frag -> raw(frag) }
    }
}
\end{lstlisting}
\end{minipage}
\end{center}

By contrast, template engines that use \textit{external} DSLs—such as JStachio,
Thymeleaf, or Handlebars—typically define templates within static HTML
documents using custom markers, and rely on blocking interfaces like
\texttt{Iterable} or \texttt{List}. These interfaces require the entire data
model, to be materialized in memory before rendering can begin, which blocks server
threads during template expansion and significantly limits scalability under
high concurrency.
Some reactive libraries, such as RxJava, provide bridging mechanisms like
\texttt{Observable.blockingIterable()}, which allows asynchronous data sources
to be exposed as \texttt{Iterable} by blocking the thread until new items are available.
While useful for compatibility with traditional APIs, this approach
reintroduces blocking behavior and undermines the benefits of non-blocking
I/O—especially under high concurrency. \autoref{lst:presentation-jstachio}
illustrates this model using a JStachio template, where the engine performs a
blocking loop over \texttt{presentationItems}.

\lstset{style=listingstyle}
\begin{center}
\begin{minipage}{0.70\textwidth}
\begin{lstlisting}[
  language=Kotlin,
  numbers=none,
  basicstyle=\scriptsize\ttfamily,
  caption={Presentation HTML template using \textit{JStachio}},
  label={lst:presentation-jstachio}
]
{{#presentationItems}}
<div class="card mb-3 shadow-sm rounded">
    <div class="card-header">
        <h5 class="card-title">
            {{title}} - {{speakerName}}
        </h5>
    </div>
    <div class="card-body">
        {{summary}}
    </div>
</div>
{{/presentationItems}}
\end{lstlisting}
\end{minipage}
\end{center}

@@

These strengths make external DSLs appealing—even when they come at the cost of
blocking, synchronous rendering. However, this tradeoff becomes critical under
high concurrency, where blocking threads severely degrades throughput~\cite{PSSR-WISE2024}.
Emerging features in the Java ecosystem, particularly \textit{virtual threads}
introduced in Java 21 as part of Project Loom~\cite{Veen2024}, offer a promising solution to
this challenge. Virtual threads drastically reduce the overhead of blocking
operations by decoupling thread execution from OS-level threads. As a result,
engines that rely on blocking interfaces—like those used in
external DSLs—can potentially achieve scalability levels that approach those of
non-blocking, asynchronous engines.

>>>>

In this section, we examine the challenges of implementing \textit{Progressive
Server-Side Rendering} (PSSR) in modern web applications, with a focus on the
limitations of current template engine designs. Our goal is to broaden the
range of options available for PSSR, particularly within JVM-based frameworks.
% Reactive streams~\cite{ReactiveStreams}, such as those based on
% \texttt{Observable} (from RxJava)~\cite{rx-observable} and \texttt{Flow} (from
% the Java 9+ standard library~\footnote{~\url{https://openjdk.org/jeps/266}}),
% are abstractions that provide a non-blocking, asynchronous
% way to handle data streams.
% These abstractions support the incremental generation and transmission of HTML
% content as data becomes available, enabling \textit{progressive server-side
% rendering} (PSSR)~\cite{pssr2005}.
In PSSR, the server does not
wait for the entire data model to be ready before beginning to render HTML.
Instead, it processes and streams each piece of data to the client as soon as
it arrives.

Reactive types like \texttt{Observable<T>} of ReactiveX~\cite{rxjava} or Kotlin \texttt{Flow<T>}~\cite{kotlinlang}
facilitate this by representing data as a sequence of asynchronous events.
For example, a reactive stream might emit a sequence of \texttt{Presentation}
objects, each representing a talk in a conference schedule.
As each \texttt{Presentation} is emitted by the \texttt{Observable}
or \texttt{Flow}, the internal DSL-based engine—such as HtmlFlow—can render the
corresponding HTML fragment and immediately flush it to the client. This
approach is demonstrated in \autoref{lst:presentation-observable}, where each
presentation is rendered asynchronously as it is emitted by an
\texttt{Observable}.
Note that the \texttt{await} builder receives an additional parameter, the
\texttt{onCompletion} callback, which is used to signal HtmlFlow that it can
proceed to render the next HTML element in the web
template~\cite{carvalho2023async}.
HtmlFlow pauses the rendering process until \texttt{onCompletion} is called,
similar to how the \emph{resume} function works in continuations and
coroutines~\cite{coroutines_continuations}.
In \autoref{lst:presentation-flow}, we show an
equivalent suspend-based implementation using Kotlin's \texttt{Flow}~\cite{wise2024pssr}.
Both examples highlight how internal DSLs can natively integrate with reactive types
to enable non-blocking, progressive rendering on the server side.

\begin{center}
\begin{minipage}{0.50\textwidth}
\begin{lstlisting}[
    language=Kotlin,
    basicstyle=\scriptsize\ttfamily,
    numbers=none,
    caption={\textit{HtmlFlow} presentation template in Koltin with an \texttt{Observable} model},
    label={lst:presentation-observable}
]
await { div, model, onCompletion ->
  model
    .doOnNext { presentation ->
      presentationFragmentAsync
        .renderAsync(presentation)
        .thenApply { frag -> div.raw(frag) }
      }
    .doOnComplete { onCompletion.finish() }
    .subscribe()
}
\end{lstlisting}
\end{minipage}
\hfill
\begin{minipage}{0.46\textwidth}
\begin{lstlisting}[
    language=Kotlin,
    basicstyle=\scriptsize\ttfamily,
    numbers=none,
    label={lst:presentation-flow},
    caption={\textit{HtmlFlow} presentation template in Kotlin with a \texttt{Flow} model},
]
suspending { model ->
  model
    .toFlowable()
    .asFlow()
    .collect { presentation ->
      presentationFragmentAsync
        .renderAsync(presentation)
        .thenApply { frag -> raw(frag) }
    }
}
\end{lstlisting}
\end{minipage}
\end{center}

By contrast, template engines that use \textit{external} DSLs—such as JStachio,
Thymeleaf, or Handlebars—typically define templates within static HTML
documents using custom markers, and rely on blocking interfaces like
\texttt{java.util.Iterable} or \texttt{java.util.stream.Stream}.
These interfaces require the entire data
model, to be materialized in memory before rendering can begin, which blocks server
threads during template expansion and significantly limits scalability under
high concurrency.
Some reactive libraries, such as RxJava, provide bridging mechanisms like
\texttt{Observable.blockingIterable()}, which allows asynchronous data sources
to be exposed as \texttt{Iterable} by blocking the thread until all items are available.
While useful for compatibility with traditional APIs, this approach
reintroduces blocking behavior and undermines the benefits of non-blocking
I/O—especially under high concurrency. \autoref{lst:presentation-jstachio}
illustrates this model using a JStachio template, where the engine performs a
blocking loop over \texttt{presentationItems}.

\begin{center}
\begin{minipage}{0.70\textwidth}
\begin{lstlisting}[
  language=Kotlin,
  numbers=none,
  basicstyle=\scriptsize\ttfamily,
  caption={Presentation HTML template using \textit{JStachio}},
  label={lst:presentation-jstachio}
]
{{#presentationItems}}
<div class="card mb-3 shadow-sm rounded">
    <div class="card-header">
        <h5 class="card-title">
            {{title}} - {{speakerName}}
        </h5>
    </div>
    <div class="card-body">
        {{summary}}
    </div>
</div>
{{/presentationItems}}
\end{lstlisting}
\end{minipage}
\end{center}

@@

These strengths make external DSLs appealing—even when they come at the cost of
blocking, synchronous rendering. However, this tradeoff becomes critical under
high concurrency, where blocking threads severely degrades throughput~\cite{wise2024pssr}.
Emerging features in the Java ecosystem, particularly \textit{virtual threads}
introduced in Java 21 as part of Project Loom~\cite{Veen2024}, offer a promising solution to
this challenge. Virtual threads drastically reduce the overhead of blocking
operations by decoupling thread execution from OS-level threads. As a result,
engines that rely on blocking interfaces—like those used in
external DSLs—can potentially achieve scalability levels that approach those of
non-blocking, asynchronous engines.

### 2025-05-27 docs/article/Sections/state-of-the-art-web-templates.tex

     & \shortstack{\texttt{Iterable} \\\texttt{Stream} \\\texttt{Publisher}}

@@

     & \shortstack{\texttt{Iterable} \\\texttt{Sequence}}

@@

     & \textcolor{PineGreen}{\checkmark}$^{(2)}$

@@

    \\$^{(1)}$ Limited to a single model per web template and compatible with
    the reactive stream \texttt{Publisher}.
    \\$^{(2)}$ Non-safety for HTML attributes.

@@

The first half of Table~\ref{table:cmplibs} lists template engines that use
their own templating dialects, referred to as \textit{external DSLs}. The
second half lists Java libraries that provide an \textit{internal DSL},
typically using a \textit{nested} or \textit{chaining} style to build HTML.
Note that, by leveraging the host language (e.g., Clojure, Groovy, Java,
Kotlin, or Scala) as the templating idiom, the latter impose no restrictions on
the data model and fully support all styles of data access APIs. They also
benefit from static type checking, which helps ensure type safety.

@@

\subsubsection{Templating idiom}

Web templates are based on a \textit{domain-specific language}
(DSL)~\cite{landin1966next}, which defines a language tailored to a specific
\textit{domain}~\cite{evans2004domain}—in this case, HTML for expressing web
documents. The DSL constrains the template's syntax and semantics to match the
structure and purpose of HTML.

DSLs can be divided in two types: \textit{external} or
\textit{internal}\cite{dslbook}. \textit{External} DSLs are languages created
without any affiliation to a concrete programming language. An example of an
\textit{external} DSL is the regular expressions search
pattern\cite{thompson1968}, since it defines its own syntax without any
dependency of programming languages. On the other hand an \textit{internal} DSL
is defined within a host programming language as a library and tends to be
limited to the syntax of the host language, such as Java.
JQuery\cite{resig2007pro} is one of the most well-known examples of an internal
DSL in Javascript, designed to simplify HTML DOM\cite{dom} tree traversal and
manipulation.

Traditionally, web template technologies use an \textit{external} DSL to define
control flow constructs and data binding primitives. Early web template engines
such as JSP, ASP, Velocity, PHP, and others adopted this \textit{external} DSL
approach as their templating dialect. For example, a \texttt{foreach} loop can
be expressed in each technology using its own DSL: \\\texttt{<\% for(String
  item : items) \%>} in JSP, \texttt{<\% For Each item In items \%>} in legacy
ASP with VBScript, \texttt{\#foreach(\$item in \$items)} in Velocity, or
\texttt{<?php foreach (\$items as \$item):?>} in PHP.

On the other hand, \textit{internal} DSLs for HTML allow templates to be
defined directly within the \emph{host} language (such as Java, Kotlin, Groovy,
Scala, or other general-purpose programming languages), rather than using
text-based template files~\cite{carvalho2019hot}. In this case, a web template
is not limited to templating constructs but may instead leverage any available
primitive of the host language or any external API.

@@

@@

One of the reasons for legacy web templates not supporting asynchronous APIs is
the absence of a unified standard calling convention for asynchronous calls.
While there is a single, straightforward way to use a synchronous API with a
direct style, where the result of a method call corresponds to its returned
value, there is no equivalent standard in the asynchronous approach. Instead,
we may encounter various asynchronous conventions depending on the programming
language and runtime environment. Some of these approaches include
\emph{continuation-passing style} (CPS)~\cite{scheme},
\textit{promises}~\cite{promise} async/await idiom~\cite{async_await}, reactive
streams \cite{ReactiveStreams}, Kotlin Flow~\cite{kotlinlang}, and others.

Many \textit{general-purpose languages} (GPLs) have embraced the
\texttt{async}/\texttt{await} feature~\cite{async_await} enabling non-blocking
routines to mimic the structure of synchronous ones, allowing developers to
reason about instruction flow sequentially. The simplicity and broad adoption
of this programming model have led to its incorporation into mainstream
languages like C\#, JavaScript, Python, Perl, Swift, Kotlin, and others,
excluding Java. However, implementing \texttt{async}/\texttt{a}wait requires
compiler support to translate \textit{suspension points} (i.e., \texttt{await}
statements) into state machines. Most template engines operate using an
external DSL with their own templating dialect (e.g., Thymeleaf, JSP, Jade,
Handlebars, and others), which do not inherently leverage asynchronous
capabilities from their host GPLs.

@@

\subsubsection{Type safety and HTML safety}

Another key characteristic of such DSLs is \textbf{HTML safety}, which refers
to whether they produce only valid HTML conforming to a well-formed document.
To ensure HTML safety, a DSL API should only allow combining calls to builders
that result in valid HTML. However, most HTML DSLs using the \textit{function
  sequence} or \textit{nested function} approach cannot ensure HTML safety at
compile time.

>>>>

     & \shortstack{\texttt{Iterable} \\\texttt{Stream} \\\texttt{Publisher}$^{(1)}$}

@@

     & \shortstack{\texttt{Iterable} \\\texttt{Sequence}$^{(2)}$}

@@

     & \textcolor{PineGreen}{\checkmark}$^{(3)}$

@@

    \\$^{(1)}$ \texttt{Publisher} of Reactive Streams~\cite{ReactiveStreams}.
    Limited to a single model per web template.
    \\$^{(2)}$ \texttt{kotlin.sequences.Sequence}
    \\$^{(3)}$ Non-safety for HTML attributes.

@@

The first half of Table~\ref{table:cmplibs} lists template engines that use
their own templating dialects, referred to as \textit{external DSLs}. The
second half lists Java libraries that provide an \textit{internal DSL},
typically using a \textit{nested} or \textit{chaining} style to build HTML.
Note that, by leveraging the host language (e.g., Clojure, Groovy, Java,
Kotlin, or Scala) as the templating idiom, the latter impose no restrictions on
the data model and fully support all styles of data access APIs. They also
benefit from static type checking, which helps ensure \textit{type safety}.

@@

\subsubsection{Domain-specific language idiom}

Web templates are based on a \textit{domain-specific language}
(DSL)~\cite{landin1966next}, which defines a language tailored to a specific
\textit{domain}~\cite{evans2004domain}—in this case, HTML for expressing web
documents. The DSL constrains the template's syntax and semantics to match the
structure and purpose of HTML.
DSLs can be divided in two types: \textit{external} or
\textit{internal}\cite{dslbook}. \textit{External} DSLs are languages created
without any affiliation to a concrete programming language. An example of an
\textit{external} DSL is the regular expressions search
pattern\cite{thompson1968}, since it defines its own syntax without any
dependency of programming languages. On the other hand an \textit{internal} DSL
is defined within a host programming language as a library and tends to be
limited to the syntax of the host language, such as Java.
JQuery\cite{resig2007pro} is one of the most well-known examples of an internal
DSL in Javascript, designed to simplify HTML DOM\cite{dom} tree traversal and
manipulation.

Traditionally, web template technologies use an \textit{external} DSL to define
control flow constructs and data binding primitives. Early web template engines
such as JSP, ASP, Velocity, PHP, and others adopted this \textit{external} DSL
approach as their templating dialect. For example, a \texttt{foreach} loop can
be expressed in each technology using its own DSL:
\begin{itemize}
  \item \texttt{<\% for(String item : items) \%>} in JSP
  \item \texttt{<\% For Each item In items \%>} in legacy ASP with VBScript
  \item \texttt{\#foreach(\$item in \$items)} in Velocity
  \item \texttt{<?php foreach (\$items as \$item):?>} in PHP
\end{itemize}

On the other hand, \textit{internal} DSLs for HTML allow templates to be
defined directly within the \emph{host} language (such as Java, Kotlin, Groovy,
Scala, or other general-purpose programming languages), rather than using
text-based template files~\cite{carvalho2020}. In this case, a web template
is not limited to templating constructs but may instead leverage any available
primitive of the host language or any external API.

@@

For example, Thymeleaf supports both Java \texttt{Iterable} and \texttt{Stream}
iteration protocols but is incompatible with Kotlin's \texttt{Sequence}
interface.
Velocity, on the other hand, uses internal reflection-based template processing
that allows it to iterate over any type defining an \texttt{iterator()} method
returning an \texttt{Iterator}. This enables support for both Java
\texttt{Iterable} and Kotlin \texttt{Sequence}, but it cannot handle Java
\texttt{Stream} since \texttt{Stream} does not expose a compatible
\texttt{iterator()} method.
The remaining analyzed template engines that use external DSLs support only Java
\texttt{Iterable} as their sole iteration protocol.

@@

One of the reasons for legacy web templates not supporting asynchronous APIs is
the absence of a unified standard calling convention for asynchronous calls.
While there is a single, straightforward way to use a synchronous API with a
direct style, where the result of a method call corresponds to its returned
value, there is no equivalent standard in the asynchronous approach. Instead,
we may encounter various asynchronous conventions depending on the programming
language and runtime environment. Some of these approaches include
\emph{continuation-passing style} (CPS)~\cite{scheme},
\textit{promises}~\cite{promise}, async/await idiom~\cite{async_await},
or \textit{suspend functions}~\cite{elizarov2021coroutines}.

Many \textit{general-purpose languages} (GPLs) have embraced the
\texttt{async}/\texttt{await} feature~\cite{async_await} enabling non-blocking
routines to mimic the structure of synchronous ones, allowing developers to
reason about instruction flow sequentially. The simplicity and broad adoption
of this programming model have led to its incorporation into mainstream
languages like C\#, JavaScript, Python, Perl, Swift, Kotlin, and others,
excluding Java. However, implementing \texttt{async}/\texttt{a}wait requires
compiler support to translate \textit{suspension points} (i.e., \texttt{await}
statements) into state machines. Most template engines operate using an
external DSL with their own templating dialect (e.g., Thymeleaf, JSP, Jade,
Handlebars, and others), which do not inherently leverage asynchronous
capabilities from their host programming languages.

An alternative approach to asynchronous programming involves handling data as
a sequence of asynchronous events, commonly known as a \textit{reactive stream}.
Instead of materializing all items in memory, reactive streams emit values over
time, allowing clients to subscribe and consume data incrementally as it becomes
available.
This idea was first proposed by Meijer~\cite{rx-observable} in the .NET
framework, leading to the development of \textit{Reactive Extensions} (Rx).
Subsequently, several alternative implementations emerged in the Java ecosystem,
such as Rx Java~\cite{rxjava}, Project Reactor~\cite{projectreactor}, or the
Akka Streams~\cite{akka} library.
Later, due to the proliferation of alternative implementations, the need arose to
create a common interface to ensure interoperability between different reactive
stream libraries. This led to the formation of a working group that included
engineers from notable companies such as Lightbend, Pivotal, and Netflix.
The group began defining the \textit{Reactive Streams}
specification~\cite{ReactiveStreams}, which established a standard for
asynchronous stream processing using non-blocking, backpressure-aware data
pipelines.
The more recent Mutiny~\cite{mutiny2021}, while conforming to the
\textit{Reactive Streams} standard, proposes a different approach that avoids
complex optimizations, resulting in more straightforward and easier-to-maintain
code.

Thymeleaf was one of the first web template engines—and remains the only one
among external DSLs for HTML—to support reactive streams based on the
\texttt{Publisher} model. However, it is limited to using a single
\texttt{Publisher} within a given template.

@@

\subsubsection{HTML safety}

Another key characteristic of such DSLs is \textit{HTML safety}, which refers
to whether they produce only valid HTML conforming to a well-formed document.
To ensure HTML safety, a DSL API should only allow combining calls to builders
that result in valid HTML. However, most HTML DSLs using the \textit{function
  sequence} or \textit{nested function} approach cannot ensure HTML safety at
compile time.

### 2025-05-27 docs/article/Sections/state-of-the-art.tex

In this section, we first examine the different design choices adopted by web
servers in their internal architectures, and how these choices impact the
behavior of the web template engines. Then, in Subsection~2.2, we present the
main properties that characterize each web template technology approach, along
with the advantages and drawbacks that result from these characteristics.

\input{Sections/state-of-the-art-web-servers.tex}
\input{Sections/state-of-the-art-web-templates.tex}

>>>>

In this section, we first present the main properties that characterize each web
template technology approach, along with the advantages and drawbacks that result
from these characteristics. Then, in Subsection~2.2, we examine the different
design choices adopted by web servers in their internal architectures, and how
these choices impact the behavior of the web template engines.

\input{Sections/state-of-the-art-web-templates.tex}
\input{Sections/state-of-the-art-web-servers.tex}
